---
title: "Object Oriented Programming in R and Python"
output: html_document
---

# Introduction

In most of statistical programming, scientists work with *numerical* and *text* data. These are useful for accomplishing typical data analysis tasks, but this is not how programming itself was designed.

Programming languages in general, are designed to give their users the ability to instruct the computer to execute any command they can think of -- not just those in the documentation. You *technically could* design an entire web server, building a self-driving car, or communicate with a satellite, all from within R. But none of these tasks are within R's strengths, and that's why other languages exist.

In general though, most mid- to low-level programming languages give you the ability to accomplish a myriad of tasks by leveraging **Object Oriented Programming**.

> Object-oriented programming (OOP) refers to a type of computer programming (software design) in which programmers define the data type of a data structure, and also the types of operations (functions) that can be applied to the data structure.

Instead of being confined to primitive datatypes -- strings, numeric, vectors, matrices, etc. --  OOP allows you as a programmer to wrap up and combine these data structures into something that works more specifically for your desired application.

# Why Should I Care?

If you've been getting along fine all this time without OOP, why should you care about this tutorial?

The simplest advantage is that OOP gives you the ability to be far more creative in solving problems. Not only will you learn how to create your own data structures for your specific use cases, you will also learn how *others* have created their data structures and the possibilities/limitations they place in their paradigms, meaning you'll be able to use their software more effectively.

If anything, this will hopefully be useful computer science knowledge in your back pocket that you can implement in your own science if you ever need to.

# Disclaimer

I'm not an expert in OOP; this only applies to the two languages I have solid experience in. OOP in other languages sometimes works quite differently, but since I work with scientists, these are the approaches I've found to be useful for scientific programming.

# OOP Theory

> Classes: A reusable collection of data and behaviour

Classes are programmers' conceptualisations of real things -- a blueprint, if you will. Instead of being forced to represent a person as a row of data on a table with columns representing attributes, you could represent them as a datatype called `person`.

A programmer can create an `instance` of a `person`, by saying that a person is an object (whoops) with a **name**, and an **age**.

What's interesting is that these pieces of data, in OOP called `attributes`, are themselves primitive datatypes, `string` and `numeric`.

Additionally, the programmer can define a finite set of actions that a `person` can do, such as `introduce` themselves, or `catch` a ball. These actions are called `methods`.

You could imagine that every instance of `person` has unique attributes for `name` and `age`, but surely every `person` should have the same species? In that case, you can define your class `person` to ALWAYS have the same species by setting it as a *class attribute*.

What's interesting, is that classes in OOP can *inherit* from one another, meaning that you could create two subclasses `male` and `female`, derived from the `person` class. The only difference being that the `male` and `female` classes have the `sex` attribute. We call `person` the 

---

### tl;dr --> Classes are blueprints for datatypes; they have attributes, that describe *what they are*, and methods, that describe *what they do*.

---

# OOP Example 1: Python

The best way to learn OOP is through a language like `Python`. It implements the most intuitive OOP paradigm and its usefulness scales with practice very smoothly. For now, we will simply demonstrate the above in `Python` before demonstrating it in `R`.

Note that this is an RMarkdown document, so we use the `reticulate` package to give us a Python engine for RMarkdown; basically, you can call python in RStudio in your RMarkdown document by changing the chunk engine from `{r}` to `{python}`:

```{r}
library(reticulate)
use_python("/usr/bin/python3")
```

```{python}
import sys
print(sys.version)
print("hello from python")
```

## How do you define a class?

First, you need to have a special python function called an *initialiser*. This function tells the class **how to set itself up**:

```{python}
class Person:
    
    # Initializer to create Attributes
    def __init__(self, name, age):
        self.name = name
        self.age = age
```

Don't get tripped up by the `self` keywords; they are basically there to tell your new object that this is a function that is self-referential. Additionally, the underscores simply help python to reserve this function and not confuse it with others written outside of the class.

Now that we have the initialiser, we can create a new instance of `Person`!

```{python}
# define a person
class Person:
    
    # Initializer to create Attributes
    def __init__(self, name, age):
        self.name = name
        self.age = age

# instantiate a person
tinashe = Person("Tinashe", 25)

# introduce myself
print("My name is ", tinashe.name, " and I'm ", tinashe.age)
```

```{python}
### Use this space to define and create a `Person` object of yourself!
### Additionally, in your class definition, add an attribute for your favourite colour (a string variable)
```

Notice that we use the `.` to access the attributes of the new `Person` object. You see this a lot in python code because many developers use OOP to create objects for all sorts of applications. To list the attributes of an object, you can use the function `dir()`:

```{python}
# define a person
class Person:
    
    # Initializer to create Attributes
    def __init__(self, name, age):
        self.name = name
        self.age = age

# instantiate a person
tinashe = Person("Tinashe", 25)

# what is a person object made of?
print(dir(tinashe))
```

## Giving a Class Methods

We can make this class do specific things by giving it methods. Let's get a person to `introduce` themselves:

```{python}

# define a person
class Person:
    
    # Initializer to create Attributes
    def __init__(self, name, age):
        self.name = name
        self.age = age
        
    def introduce(self):
        return "My name is {} and I'm {} years old".format(self.name, self.age)

# instantiate a person
tinashe = Person("Tinashe", 25)

print(tinashe.introduce())
        
```

Again, note the use of `self`. By defining the function in the class with `self` as the first argument, it allows the function to access and manipulate the attributes of the object. We can create functions that do that to our `person`:

```{python}

# define a person
class Person:
    
    # Initializer to create Attributes
    def __init__(self, name, age):
        self.name = name
        self.age = age
        self.hair = None
        
    def introduce(self):
        if self.hair is None:
            return "My name is {} and I'm {} years old".format(self.name, self.age)
        else:
            return "My name is {}, I'm {} years old, and my hair colour is {}".format(self.name, self.age, self.hair)

    def set_hair_colour(self, colour):
        self.hair = colour

# instantiate a person
tinashe = Person("Tinashe", 25)

# before we set hair colour
print tinashe.introduce()

# after we set hair colour
tinashe.set_hair_colour("Black")
print(tinashe.introduce())
```

Pretty cool!

```{python}
### Use this space to create a person class with a method to tell you how old they'll be in x years.
### Hint: use `def` and don't forget to use `self`

### def future_age(self, x):
###     current_age = self.age
###     ...
###     ...
###     ...
```


## Class inheritance

An important understanding for OOP is class inheritance. Just as a `student` is a type of `Person`, classes in python can be subclasses of other classes. Let's create a student:

```{python, eval = FALSE}

# define a person
class Person:

    # Initializer to create Attributes
    def __init__(self, name, age):
        self.name = name
        self.age = age
        self.hair = None

    def introduce(self):
        if self.hair is None:
            return "My name is {} and I'm {} years old".format(self.name, self.age)
        else:
            return "My name is {}, I'm {} years old, and my hair colour is {}".format(self.name, self.age, self.hair)

    def set_hair_colour(self, colour):
        self.hair = colour

# define a student
class Student(Person):


    # we need to instantiate it similarly to the person
    def __init__(self, name, age, field):
        self.field = field
        super().__init__(name, age)      # this is a special way of saying "do everything else for the parent class as normal"

    def field_of_study(self):
        return "I study {}".format(self.field)

# instantiate a student, using the same arguments as Person
tinashe = Student("Tinashe", 25, "neuroscience")

print(tinashe.introduce())
print(tinashe.field_of_study())
```

Hopefully this gives you a grand idea of what OOP is and how it works. Developers use OOP to implement real-world ideas in programming languages. With the correct blueprint and enough cleverness, you could model the diffusion of point particles, a taxonomy of animals, products in a supply chain, etc. etc.

### tl;dr --> We learned how to create classes in Python, by creating objects with attributes, and methods that we can use to access/manipulate these attributes; we also learned how to create a subclass!

---

# OOP Example 2: R

We start with Python because it's the most intuitive and user friendly for OOP insofar as understanding *what it is*. In R, this is somewhat clunkier and opaque -- with good reason! Most of us in R only care about dataframes and the values in their cells.

What you may not realise is that R has a lot of class programming going on. A `tibble`, for example, is a subclass of `data.frame`, meaning that everything a `data.frame` can do, a `tibble` can do as well. The reason a `tibble` prints differently from a `data.frame` is that the clever folks at RStudio used OOP to **override** the print method of a `data.frame` and make it prettier. 

```{r}
print(iris)
print(tibble::as_tibble(iris))
```

Understanding how these OOP principles work in any language is valuable for solving programming problems -- both in creating your own classes and using others' classes. In R, developers use OOP to take advantages of **polymorphism** and **encapsulation**:

> Polymorphism means that a developer can consider a function’s interface separately from its implementation, making it possible to use the same function form for different types of input. This is closely related to the idea of encapsulation: the user doesn’t need to worry about details of an object because they are encapsulated behind a standard interface.

To be concrete, polymorphism is what allows summary() to produce different outputs for numeric and factor variables:

```{r}

diamonds <- ggplot2::diamonds

summary(diamonds$carat)
summary(diamonds$cut)
```

What's happening is that the class of the `diamonds$carat` vector is defined differently from the `diamonds$cut` vector. Calling `summary` on these produces different results because `summary` looks at the class of the object it's being called on first, and then *dispatches* the correct method for that class. That's why you sometimes see, in documentation:

```
summary
summary.aov
summary.glm
summary.lm
etc.
```
`summary` is the interface for a function that acts differently depending on the class. Let's learn how to create classes in R and override the summary function accordingly. In R, we call these functions `generics` and these perform the job of

## Simple OOP: S3

**S3** refers to the OOP paradigm first designed for S, version 3 [see WhyDoesR](https://twitter.com/whydoesr). It comes in Base and is very simple/flexible. It's not nearly as strict as the python paradigm -- all you have to do is set the `class` attribute:

```{r}
# first, create a named vector
tinashe <- c("Tinashe", 25)
names(tinashe) <- c("Name", "Age")

#set the class attribute
class(tinashe) <- "Person"

tinashe
```

That's it!

In practice though, for S3, you should try to emulate the behaviours we practiced in Python for safety:

```{r}
# a constructor; in python we called this the initialiser in the class blueprint

new_Person <- function(name, age){
  
  if(typeof(name) != "character") {
    stop("Name must be a character vector")
  }
  
  if(typeof(age) != "double") {
    stop("Age must be a numeric")
  }
  
  person <- c(name, age)
  names(person) <- c("name", "age")
  
  class(person) <- "Person"
  
  person

}
```

## Method Dispatch

A bit confusingly (WhyDoesR?!), in order to define a method for a class, you need to understand method dispatch. Method dispatch is a type of strategy for passing a class around until it finds the right method. E.g., `summary()` dispatches the class to the correct method depending on the object it has to summarise. 
